On Optimality of Finite Element Integration

### 2015-09-24 TOMS_Paper/assembly.tex
Several fixes, update plots, descriptions, etc

Experiments were run on a single core of an Intel I7-2600 (Sandy Bridge) CPU, running at 3.4GHz, 32KB L1 cache (private), 256KB L2 cache (private) and 8MB L3 cache (shared). The Intel Turbo Boost and Intel Speed Step technologies were disabled. The Intel \texttt{icc 15.2} compiler was used. The compilation flags used were \texttt{-O3, -xHost, -ip}, which can trigger \texttt{AVX} autovectorization.

We analyze the runtime performance in four real-world bilinear forms of increasing complexity, which comprise the differential operators that are most common in finite element methods. In particular, we study the mass matrix (``\texttt{Mass}'') and the bilinear forms arising in a Helmholtz equation (``\texttt{Helmholtz}''), in an elastic model (``\texttt{Elasticity}''), and in a hyperelastic model (``\texttt{Hyperelasticity}''). The complete specification of these forms is made publicly available\footnote{\url{https://github.com/firedrakeproject/firedrake-bench/blob/experiments/forms/firedrake_forms.py}}.

We evaluate the speed-ups achieved by a wide variety of transformation systems over the original code (i.e., no optimizations applied) as returned by the FEniCS Form Compiler. We analyze the impact of
\begin{itemize}
\item FEniCS Form Compiler: optimized quadrature mode (work presented in~\citeN{quadrature1}). Referred to as \texttt{quad} 
\item FEniCS Form Compiler: tensor mode (work presented in~\citeN{FFC-TC}). Referred to as \texttt{tens} 
\item FEniCS Form Compiler: automatic mode (choice between \texttt{tens} and \texttt{quad} driven by heuristic, detailed in~\citeN{Fenics} and summarized in Section~\ref{sec:qualitative}). Referred to as \texttt{auto} 
\item UFLACS: a novel back-end for the FEniCS Form Compiler (whose primary goals are improved code generation time and runtime). Referred to as \texttt{ufls} 
\item COFFEE: generalized loop-invariant code motion (work presented in~\citeN{Luporini}). Referred to as \texttt{cfO1} 
\item COFFEE: optimal loop nest synthesis plus symbolic execution for zero-elimination (work of this article). Referred to as \texttt{cfO2} 
\end{itemize}

The values that we report are the average of three runs with ``warm cache'' (no code generation time, no compilation time). They include the cost of local assembly as well as the cost of matrix insertion. However, the unstructured mesh used to run the simulations was chosen small enough to fit the L3 cache of the CPU, so as to minimize the ``noise'' due to operations outside of the element matrix evaluation.

For the fairest comparison possible, small patches (publicly available) were written to be able to run all simulations through Firedrake: this means the cost of matrix insertion and mesh iteration is exactly the same in all variants. UFLACS and the FEniCS Form Compiler's optimization systems generate code suitable for FEniCS, which employs a data storage layout different than Firedrake. Our patches fix this problem by producing code with a data storage layout as expected by Firedrake.

For what concerns the second memory constraint introduced in Section~\ref{sec:mem-const}, we set $T_H = L2_{size}$; that is, the amount of space that temporaries due to pre-evaluation can occupy is bounded by the size of the processor L2 cache (the last level of private cache). For the test cases in which $T_H$ was determinant to prevent pre-evaluation, we also repeated the experiments setting $T_H = L3_{size}$ to assess our hypotheses. Results of these trials are discussed below.

Following the methodology adopted in~\citeN{quadrature1}, we increasingly vary the complexity of each form. In particular:
\begin{itemize}
\item the polynomial order of test and trial functions, $q \in \lbrace1, 2, 3, 4\rbrace$. Test, trial, and coefficient (or ``pre-multiplying'') functions always have same degree
\item the number of coefficient functions $nf \in \lbrace0, 1, 2, 3\rbrace$
\end{itemize}
Constants of our analysis, instead, are
\begin{itemize}
\item the space of test, trial, and coefficient functions: Lagrange
\item the mesh: tetrahedral with a total of 4374 elements
\item exact numerical quadrature (the same scheme as in~\citeN{quadrature1}, based on the Gauss-Legendre-Jacobi rule, is employed)
\end{itemize}

The upcoming Figures (\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity}) are three-dimensional plots. The three axes are for $q$, $nf$, and code transformation system. The code transformation system varies within different subplots, such that one can easily evince the fastest alternative for each problem instance: the taller a bar, the better, since, as explained, we are reporting speed-ups.

%TODO : can we just say padding is useless because in 3D ndofs is a multiple of our vector length?

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/mass}}
\caption{Performance evaluation for the mass matrix. The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:mass}
\end{figure}

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/helmholtz}}
\caption{Performance evaluation for the bilinear form of a Helmholtz equation. The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:helmholtz}
\end{figure}

\begin{figure}
\includegraphics[scale=0.77]{perf-results/elasticity}
\caption{Performance evaluation for the bilinear form arising in a elastic model. The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:elasticity}
\end{figure}

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/hyperelasticity}}
\caption{Performance evaluation for the bilinear form arising in a hyperelastic model. The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:hyperelasticity}
\end{figure}

\subsection{Performance of Forms}
\label{sec:perf-results-forms}
% 52 test cases per form, so 52*4 = 208 test cases in total. In 7, there is a non-marginal slow down in cfO2.

The first observation is that our optimality model does not imply minimum execution time. In particular, in the 0.06$\%$ of the test cases (we are not counting marginal differences), \texttt{cfO2} does not result in the best runtime. There are two reasons for this. The former is that low level optimization can have a significant impact. Compare, as an example, \texttt{tens} and \texttt{cfO2} in \texttt{Mass} and \texttt{Helmholtz} when $q=1$. With $p \in [2,3]$, \texttt{tens} slightly outperforms \texttt{cfO2}. With $p=4$, \texttt{tens} is quite faster than \texttt{cfO2}. The motivation is the aggressive unrolling performed in \texttt{tens}, which 1) allows to eliminate all operations involving zero-valued entries and 2) can drastically decrease both working set and register pressure by avoiding storing entire temporaries (there are often many repeated values that, with unrolling, need be allocated only once). In COFFEE we never unroll loops to maximize the chances of autovectorization, which usually results in considerable speed-ups throughout the majority of test cases. The latter reason is simply inherent to our optimality model. As explained in Section~\ref{sec:heuristic-int-dep}, we use heuristics to optimize the integration loop. The reason we are studying an hyperelastic model is indeed to assess their effectiveness; the systematic performance improvements observed over all other variants are, however, extremely encouraging.

The second aspect regards the ``triggering'' of pre-evaluation in \texttt{cfO2}. The fact that a bar in \texttt{cfO2} matches the corresponding bar in \texttt{cfO1} usually suggests that pre-evaluation was found unprofitable. The trend -- although there are quite a few exceptions -- is that as $nf$ increases, pre-evaluation tends not to be performed. This is a consequence of a larger increase factor, as discussed in Section~\ref{sec:op_count}.

The third point concerns the effect of the chosen discretizations on autovectorization. The discretizations employed ensure both multilinear loop nests and tabulated basis function sizes to be multiple of the machine vector length. Given the suitable choice of compilation options, this promotes autovectorization in the majority of code variants. The biggest exception is \texttt{quad}, in which, due to the presence of indirection arrays in the generated code, vectorization is rarely triggered. In \texttt{tens}, the loop nests are fully unrolled, so the standard loop vectorization is not possible; however, manual inspection of the compiled code suggests block vectorization (\cite{SLP-vect}) is often triggered. In \texttt{ufls}, \texttt{cfO1}, and \texttt{cfO2} the iteration spaces tend to have the same structure (there are a few exceptions in \texttt{cfO2} due to zero-elimination), with loop vectorization being regularly applied, as far as we could understand.

Also note some other minor things.
\begin{itemize}
\item The lack of (or marginal) improvements in all code variants when $[q=1, p=1]$ is due to the cost of matrix insertion, which outweighs that of local assembly. 
\item The slow-downs experienced with \texttt{quad} are imputable to the presence of indirection arrays in the generated code; especially with dense tables, trading vectorization for avoiding iteration over a few zero-valued columns may be counter-productive (see also Section~\ref{sec:zeros}).
\item A missing bar in \texttt{tens} or \texttt{auto} means that the code generation system failed because of either exceeding the memory limit or being unable to manipulate the math characterizing the form (this happens systematically in the hyperelastic model, and for some discretizations in the elastic model).
\item The bars for $nf=0$ do not change as $p$ increases: this is expected since varying $p$ when there are no coefficients should not make any difference.
 \end{itemize}

\paragraph{Mass}
From analysis of Figure~\ref{fig:mass}, we can see that \texttt{cfO2} is the best variant among the ones tested, apart from a few points already discussed. This suggests that the optimality model holds in these simple bilinear forms; that is, \texttt{cfO2} generates loop nests that are optimal from both a theoretical and practical viewpoints. It is worthwhile noting how \texttt{auto} is not capable of selecting the proper synthesis in many of the test cases, especially as $nf \geq 2$. The fact that \texttt{cfO2} is slower than \texttt{tens} at $[q=4, p=4, nf=1]$ is explained by the memory threshold that prevents pre-evaluation. By just setting $T_H = L3_{size}$, \texttt{cfO2} becomes 1.48$\times$ faster than \texttt{tens}. This is probably due to the hardware prefetching being way more effective than in all other forms: the number of allocated temporaries is relatively small, and so is the number of memory access traces that need be tracked. However, as we already explained, in a parallel context this gain might diminish, since the L3 cache is shared by the cores of the CPU.

\paragraph{Helmholtz}
Figure~\ref{fig:helmholtz} shows that the performance improvements achieved by various code variants over a non-optimized implementation can be significant. We appreciate the fact that the relative order of the employed function spaces plays a key role in the code synthesis choice. Take, for example, the case $[q=3, p=1, nf=2]$. The adoption of pre-evaluation makes a tremendous speed-up be achievable by \texttt{cfO2} over the competitors. On the other hand, \texttt{auto} triggers \texttt{quad} (since one monomial in the form has two derivatives and two coefficients, which exceeds the threshold within which \texttt{tens} is selected), which, just like the other quadrature-based variants \texttt{ufls} and \texttt{cfO1}, is sub-optimal. The effects of the working set threshold are in general positive. Moving to $T_H = L3_{size}$ makes \texttt{cfO2} slower than \texttt{cfO1} in many cases with $nf=2$, the most significant one being a slow-down of 2.16$\times$ at $[q=2, p=2]$.

>>>>

Experiments were run on a single core of an Intel I7-2600 (Sandy Bridge) CPU, running at 3.4GHz, 32KB L1 cache (private), 256KB L2 cache (private) and 8MB L3 cache (shared). The Intel Turbo Boost and Intel Speed Step technologies were disabled. The Intel \texttt{icc 15.2} compiler was used. The compilation flags used were \texttt{-O3, -xHost, -ip}.

We analyze the runtime performance of four real-world bilinear forms of increasing complexity, which comprise the differential operators that are most common in finite element methods. In particular, we study the mass matrix (``\texttt{Mass}'') and the bilinear forms arising in a Helmholtz equation (``\texttt{Helmholtz}''), in an elastic model (``\texttt{Elasticity}''), and in a hyperelastic model (``\texttt{Hyperelasticity}''). The complete specification of these forms is made publicly available\footnote{\url{https://github.com/firedrakeproject/firedrake-bench/blob/experiments/forms/firedrake_forms.py}}.

We evaluate the speed-ups achieved by a wide variety of transformation systems over ``original code'' produced by the FEniCS Form Compiler (i.e., no optimizations applied). We analyze the following transformation systems
\begin{itemize}
\item FEniCS Form Compiler: optimized quadrature mode (work presented in~\citeN{quadrature1}). Referred to as \texttt{quad} 
\item FEniCS Form Compiler: tensor mode (work presented in~\citeN{FFC-TC}). Referred to as \texttt{tens} 
\item FEniCS Form Compiler: automatic mode (choice between \texttt{tens} and \texttt{quad} driven by heuristic, detailed in~\citeN{Fenics} and summarized in Section~\ref{sec:qualitative}). Referred to as \texttt{auto} 
\item UFLACS: a novel back-end for the FEniCS Form Compiler (whose primary goals are improved code generation time and runtime). Referred to as \texttt{ufls} 
\item COFFEE: generalized loop-invariant code motion (work presented in~\citeN{Luporini}). Referred to as \texttt{cfO1} 
\item COFFEE: optimal loop nest synthesis plus symbolic execution for zero-elimination (work of this article). Referred to as \texttt{cfO2} 
\end{itemize}

The values that we report are the average of three runs with ``warm cache'' (no code generation time, no compilation time). They include the cost of local assembly as well as the cost of matrix insertion. However, the unstructured mesh used for the simulations (details below) was chosen small enough to fit the L3 cache of the CPU so as to minimize the ``noise'' due to operations outside of the element matrix evaluation.

For a fair comparison, small patches (publicly available) were written to run \textit{all} simulations through Firedrake. This means the costs of matrix insertion and mesh iteration are identical in all variants. Our patches make UFLACS and the FEniCS Form Compiler's optimization systems generate code suitable for Firedrake, which employs a data storage layout different than that of FEniCS (e.g., array of pointers instead of pointer to pointers).

In Section~\ref{sec:mem-const}, we discussed the importance of memory constraints. We then define $T_H$ as the maximum amount of space that temporaries due to code motion can take. We set $T_H = L2_{size}$, that is, the size of the processor L2 cache (the last level of private cache). We recall that exceeding this threshold prevents the application of pre-evaluation. In our experiments, this happened in some circumstances. In such cases, experiments were repeated with $T_H = L3_{size}$ to verify the hypotheses made in Section~\ref{sec:mem-const}. We later elaborate on this.

Following the methodology adopted in~\citeN{quadrature1}, we vary the following parameters:
\begin{itemize}
\item the polynomial order of test, trial, and coefficient (or ``pre-multiplying'') functions, $q \in \lbrace1, 2, 3, 4\rbrace$
\item the number of coefficient functions $nf \in \lbrace0, 1, 2, 3\rbrace$
\end{itemize}
While constants of our study are
\begin{itemize}
\item the space of test, trial, and coefficient functions: Lagrange
\item the mesh: tetrahedral with a total of 4374 elements
\item exact numerical quadrature (we employ the same scheme used in~\citeN{quadrature1}, based on the Gauss-Legendre-Jacobi rule)
\end{itemize}

We report the results of our experiments in Figures~\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity} as three-dimensional plots. The axes represent $q$, $nf$, and code transformation system. We show one subplot for each problem instance $[q, nf]$, with the code transformation system varying within each subplot. The best variant for each problem instance is given by the tallest bar, which indicates the maximum speed-up over non-transformed code.

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/mass}}
\caption{Performance evaluation for the \textit{mass} matrix. The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:mass}
\end{figure}

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/helmholtz}}
\caption{Performance evaluation for the bilinear form of a \textit{Helmholtz} equation. The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:helmholtz}
\end{figure}

\begin{figure}
\includegraphics[scale=0.77]{perf-results/elasticity}
\caption{Performance evaluation for the bilinear form arising in an \textit{elastic} model. The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:elasticity}
\end{figure}

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/hyperelasticity}}
\caption{Performance evaluation for the bilinear form arising in a \textit{hyperelastic} model. The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:hyperelasticity}
\end{figure}

\subsection{Performance Results}
\label{sec:perf-results-forms}

The main observation is that our transformation strategy does not always guarantee minimum execution time. In particular, 5$\%$ of the test cases (3 out of 56, without counting marginal differences) show that \texttt{cfO2} was not optimal in terms of runtime. The most significant of such test cases is the elastic model with $[q=4, nf=0]$. There are two reasons for this. Firstly, low level optimization can have a significant impact on actual performance. For example, the aggressive loop unrolling in \texttt{tens} eliminates operations on zeros and reduces the working set size by not storing entire temporaries; on the other hand, preserving the loop structure can maximize the chances of autovectorization. Secondly, memory constraints are critical, particularly the transformation strategy adpopted when exceeding $T_H$. We will later thoroughly elaborate on all these aspects.

A second observation regards the use of pre-evaluation in \texttt{cfO2}. The fact that the \texttt{cfO2} and \texttt{cfO1} bars match suggests pre-evaluation was found unprofitable and, therefore, rejected. The trend -- although there are quite a few exceptions -- is that as $nf$ increases, pre-evaluation tends not to be performed. This is a consequence of a larger increase factor, as discussed in Section~\ref{sec:op_count}.

A third point concerns autovectorization. The discretizations employed result in inner loops and basis function tables of size multiple of the machine vector length. This, combined with the chosen compilation flags, promotes autovectorization in the majority of code variants. An exception is \texttt{quad} due to the presence of indirection arrays in the generated code. In \texttt{tens}, loop nests are fully unrolled, so the standard loop vectorization is not feasible; manual inspection of the compiled code suggests, however, that block vectorization (\cite{SLP-vect}) is often triggered. In \texttt{ufls}, \texttt{cfO1}, and \texttt{cfO2} the iteration spaces have similar structure (there are a few exceptions in \texttt{cfO2} due to zero-elimination), with loop vectorization being regularly applied, as far as we could evince from compiler reports and manual inspection of assembly code.

Other less relevant aspects are
\begin{itemize}
\item The lack of (or marginal) improvements in all code variants when $q=1$ is due to the cost of matrix insertion, which outweighs that of local assembly.
\item The slow-downs experienced with \texttt{quad} are imputable to the presence of indirection arrays in the generated code; especially with dense tables, trading autovectorization for a relatively small save in floating point operations can be counter-productive (see also Section~\ref{sec:zeros}).
\item If a bar or a subplot is missing, it means that the form compiler failed at generating code because of either exceeding the memory limit of the system or being unable to handle the form.
\end{itemize}

\paragraph{Mass}
We start with the simplest of the bilinear forms that we are investigating, the mass matrix. From analysis of Figure~\ref{fig:mass}, we see that \texttt{cfO2} shows the highest speed-ups. It is worth noting how \texttt{auto} does not select the fastest implementation (out of \texttt{quad} and \texttt{tens}) in some of the test cases: \texttt{auto} always opts for \texttt{tens}, while as $nf \geq 2$ \texttt{quad} tends to be preferable. On the other hand, one can infer that \texttt{cfO2} constantly makes correct decisions about pre-evaluation.

\paragraph{Helmholtz}
The main message of Figure~\ref{fig:helmholtz} is that \texttt{cfO2} still outperforms the other code generation systems. For $nf=0$, the adoption of pre-evaluation results in increasingly notable speed-ups over \texttt{cfO1} as $q$ increases; \texttt{tens} is comparable to \texttt{cfO2}, with \texttt{auto} making the right choice. With $nf=1$, \texttt{auto} still picks \texttt{tens}, although this time the choice is sub-optimal: this can indirectly be inferred from \texttt{cfO2}, which rejects pre-evaluation for any $q$. For $nf=2$ and $nf=3$, \texttt{auto} reverts to \texttt{quad}, which would theoretically be the right choice (the flop count is much lower than in \texttt{tens} or what would be produced by pre-evaluation); however, the generated code suffers from the presence of indirection arrays, which break autovectorization and ``traditional'' code motion. The sporadic slow-downs or marginal improvements exhibited by \texttt{ufls} are imputable to the presence of sharing.

An interesting experiment we performed was relaxing the memory threshold by setting it to $T_H = L3_{size}$. We found that this makes \texttt{cfO2} generally slower as $nf \geq 2$, with a maximum slow-down of 2.16$\times$ with $nf=2, q=2$. Presumably, the effects of not having a sensible threshold could even be worse in parallel runs, since the L3 cache is a shared memory.
